The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-06-07 pkg/partykit/inst/guideliketest/draft/draft.Rnw
adapted draft.Rnw to restructured replication material and deleted old results
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2617 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 3a7ac63ca25bf4009378d3b0946c1aa9f7481a68

source("../guidelike.R")
source("../tests_guideprune.R")
source("../ccprune.R")

if(file.exists("stump.rda")){
  load("stump.rda")
} else {
  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
                         nrsplitvar = 1, 
                         nrsteps = 1, nrlevels = 2,
                         delta = seq(from = 0, to = 1, by = 0.1),
                         xi = c(0, 0.2, 0.5, 0.8), 
                         vary_beta = c("all", "beta0", "beta1"),
                         binary_regressor = FALSE,
                         binary_beta = c(TRUE, FALSE),
                         only_intercept = FALSE,
                         test = c("ctree", "mfluc", "ctree_max",
                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
                       beta0 = 0, beta1 = 1,
                       stump = TRUE, z1dist = "unif", sigma = 1, alpha = 0.05)

  save(simres, file = "stump.rda")
}
sim_stump <- simres$res
rm(simres)

#if(file.exists("tree.rda")){
#  load("tree.rda")
#} else {
#  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7,
#                         nrsplitvar = 2, nrsteps = 1, nrlevels = 2,
#                         delta = seq(from = 0, to = 1, by = 0.1),
#                         xi = c(0, 0.2, 0.5, 0.8), vary_beta = "all",
#                         binary_regressor = FALSE, binary_beta = TRUE,
#                         only_intercept = FALSE,
#                         test = c("ctree", "mfluc", "ctree_max",
#                                  "ctree_cat", "mfluc_cat", "ctree_max_cat",
#                                  "ctree_bin", "mfluc_bin", "ctree_max_bin",
#                                  "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
#                                  "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"),
#                         beta0 = NULL, beta1 = NULL,
#                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05)
#  
#  save(simres, file = "tree.rda")
#}
#sim_tree <- simres$res
#rm(simres)

@@

d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)

@@

\minipage{0.45\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, binary_beta = TRUE, binary_regressor = FALSE, only_intercept = FALSE,
               vary_beta = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three variations: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

<<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
# d <- dgp_tree(nobs = 250)
set.seed(7)
d <- dgp_tree(nobs = 250, xi = 0, binary_regressor = FALSE, vary_beta = "all", 
              binary_beta = TRUE, nrsplitvar = 2, sigma = 0.5)

@@

library("latticeExtra")
useOuterStrips(xyplot(prop_Tsplit ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Selection probability of"~Z[1]),
                      par.settings = par.settings),

@@

<<3way_data, results=hide, echo=FALSE>>=
if(file.exists("3way_xi00_delta03.rda")){
  load("3way_xi00_delta03.rda")
} else {
  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
                delta = 0.3,
                xi = 0, 
                binary_regressor = FALSE,
                binary_beta = TRUE,
                only_intercept = FALSE,
                nrlevels = 2, nrsplitvar = 1,
                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
                return_matrices = TRUE,
                test = c("guide_sum_1_cor",
                         "mfluc_resid_bin",
                         "ctree_resid_bin",
                         "mfluc_resid_cat",
                         "mfluc_resid",
                         "ctree_resid",
                         "guide_sum_12",
                         "mfluc_bin",
                         "ctree_bin",
                         "mfluc_cat",
                         "mfluc",
                         "ctree"))
  d00 <- prep_3way(simres)
  save(d00, file = "3way_xi00_delta03.rda")
}

if(file.exists("3way_xi08_delta1.rda")){
  load("3way_xi08_delta1.rda")
} else {
  simres <- sim(nobs = 250, nrep = 100, seed = 7, stump = TRUE, nrsteps = 1,
                delta = 1,
                xi = 0.8, 
                binary_regressor = FALSE,
                binary_beta = TRUE,
                only_intercept = FALSE,
                nrlevels = 2, nrsplitvar = 1,
                vary_beta = "all", beta0 = 0, beta1 = 1, alpha = 0.05,
                return_matrices = TRUE,
                test = c("guide_sum_1_cor",
                         "mfluc_resid_bin",
                         "ctree_resid_bin",
                         "mfluc_resid_cat",
                         "mfluc_resid",
                         "ctree_resid",
                         "guide_sum_12",
                         "mfluc_bin",
                         "ctree_bin",
                         "mfluc_cat",
                         "mfluc",
                         "ctree"))
  d08 <- prep_3way(simres)
  save(d08, file = "3way_xi08_delta1.rda")
}

d00$xi <- 0
d08$xi <- 0.8

d3way <- rbind(d00, d08)
d3way$xi <- factor(d3way$xi, levels = c(0,0.8), labels = c(0,0.8))
@

\begin{figure}[t!]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = d3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       scales = list(alternating=FALSE),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       #index.cond = list(c(1,2,3,4)),
       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
                                   bg = "gray90"),
       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
                  columns = 2,
                  lines = list(lty = c(0,1,2,0,1,1), 
                               type = "l", lwd = 2,
                               col = c("black","black","black","black", pal[1], pal[3]))))

#par(mar = c(3,5,1,5.7))
#with(d00, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1),
#                           ylab = expression("Mean"~p~"-value for"~Z[1])))
#with(d00, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d00, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
#with(d08, interaction.plot(res_scores, interaction(cat, bin), pval_z1, ylim = c(0,1)))
#with(d08, interaction.plot(interaction(cat, bin), res_scores, pval_z1))
#with(d08, interaction.plot(cat, interaction(bin, res_scores), pval_z1))
@
\caption{\label{fig:3way} Effects of the three building blocks \textit{residuals vs.\ scores},
\textit{dichotomization}, and \textit{categorization} in the ``stump'' scenario with the intercept
and the slope parameter both varying. All possible combinations have been 
evaluated in two different settings and their performances are compared based on the 
mean $p$-values corresponding to the true split variable $Z_1$. 
Left panel: true split point at the median $0$ and with
effect size $0.3$.
Right panel: true split point at the $90$\% quantile $0.8$ and with effect 
size $1$.}
\end{figure}

@@

<<tree_all_data, echo=FALSE, results=hide>>=
if(file.exists("pruning.rda")){
  load("pruning.rda")
} else {
  simres <- simwrapper_p(nobs = 250, nrep = 100, seed = 7, 
                         nrsteps = 1, nrlevels = 2,
                         delta = seq(from = 0, to = 1, by = 0.1),
                         xi = c(0, 0.2, 0.5, 0.8), 
                         vary_beta = "all", binary_regressor = FALSE,
                         binary_beta = TRUE, only_intercept = FALSE,
                         compare_pruning = TRUE,
                         test = c("ctree", "mfluc",
                                  "guide_sum_12", 
                                  "guide_sum_1_cor"),
                         beta0 = NULL, beta1 = NULL,
                         stump = FALSE, z1dist = "unif", sigma = 1, alpha = 0.05,
                         nrsplitvar = 2)

  save(simres, file = "pruning.rda")  
}

pr <- simres$res
## extract data for xyplot comparing ARI and ARI_p
pr <- pr[,c(1,2,7,8,9,12,15)]
## extended data set with post-pruning results being represented as separated tests
prex <- pr[,c(1,2,3,4,6)]
prex <- rbind(prex,prex)

prex$test <- factor(prex$test,
                    levels = c("ctree","mfluc",
                               "guide_sum_12", "guide_sum_1_cor"),
                    labels = c("CTree","MOB",
                               "GUIDE+scores", "GUIDE"))

prex$nrsubgr[(length(pr$nrsubgr)+1) : (2*length(pr$nrsubgr))] <- pr$nrsubgr_p
prex$ari[(length(pr$ari)+1) : (2*length(pr$ari))] <- pr$ari_p
prex$pruning <- c(rep("pre", NROW(pr)), rep("post", NROW(pr)))

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<tree_all, fig=TRUE, echo=FALSE, height=5, width=8>>=
library("latticeExtra")
useOuterStrips(xyplot(ari ~ delta | xi + pruning, groups = ~ test, 
                      data = prex, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      key =  list(text = list(levels(prex$test)[1:4]),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(4,2),
                      scales=list(x=list(at=seq(1,11,2))),
                      xlab = expression(delta),
                      ylab = "Adjusted Rand Index",
                      #index.cond = list(c(2,3,1),c(2,1)),
                      par.settings = par.settings),
               strip = strip.custom(factor.levels = c(expression(xi==0), 
                                                      expression(xi==0.2), 
                                                      expression(xi==0.5), 
                                                      expression(xi==0.8)),
                                    bg = "gray90"),
               strip.left=strip.custom(factor.levels = c("Post-pruning", "Pre-pruning"),
                                   bg = "gray90"))
@
\caption{\label{fig:tree_all}
Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
``tree'' scenario, once applying pre-pruning (top row) and once post-pruning (bottom row). 
The ARI is evaluated over
increasing effect size $\delta$ with 100 replications of 250 observations per step.
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right, corresponding to the $50$\%, $60$\%, $75$\%, and $90$\% quantile).
}
\end{figure}

@@

library("latticeExtra")
useOuterStrips(xyplot(prop_z1 ~ delta | vary_beta + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = binary_beta == TRUE & binary_regressor == FALSE & 
                        only_intercept == FALSE,# & xi == 0, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
                      par.settings = par.settings),

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
#subdata <- subset(subdata, delta %in% c(0, 0.1, 0.2, 0.3, 0.4, 0.5))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"),
                       labels = c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores"))
par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
                                             fill = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]), 
                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]))

>>>>

source("../simulation/guidelike.R")
source("../simulation/tests_guideprune.R")
source("../simulation/ccprune.R")

if(file.exists("../simulation/results/sim_stump.rda")){
  load("../simulation/results/sim_stump.rda")
} else {
  sim_stump <- simwrapper(scenario = "stump",
                          nobs = 250, nrep = 100, seed = 7,
                          delta = seq(from = 0, to = 1, by = 0.1),
                          xi = c(0, 0.2, 0.5, 0.8), 
                          variation = c("all", "beta0", "beta1"),
                          changetype = c("abrupt", "continuous"),
                          test = c("ctree", "mfluc", "ctree_max",
                                   "ctree_cat", "mfluc_cat", "ctree_max_cat",
                                   "ctree_bin", "mfluc_bin", "ctree_max_bin",
                                   "ctree_cat_bin", "mfluc_cat_bin", "ctree_max_cat_bin",
                                   "guide_sum_12", "guide_coin_12", "guide_sum_1_cor"))

  sim_stump <- sim_stump$res

  save(sim_stump, file = "../simulation/results/sim_stump")
}

@@

d <- dgp_stump(nobs = nobs, changetype= "abrupt",
               variation = "beta0", beta1 = beta1, delta = delta, xi = xi, sigma = sigma)

@@

\minipage{0.45\textwidth}
<<dgp_stump3, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, changetype= "abrupt",
               variation = "beta1", beta0 = beta0, delta = delta, xi = xi, sigma = sigma)

par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x", main = expression("varying"~beta[1]),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(0,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(0,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>xi], ylim = c(-5,5), pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -2)
mtext(expression(beta[0]==0), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 2)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1)
@
\endminipage
\minipage{0.45\textwidth}
<<dgp_stump4, fig=TRUE, echo=FALSE, width=4, height=2.9>>=
d <- dgp_stump(nobs = nobs, changetype= "abrupt", 
               variation = "all", delta = delta, xi = xi, sigma = sigma)
par(mar = c(4,4,2,4))
plot(x = d$x[d$z1<=xi], y = d$y[d$z1<=xi], ylim = c(-5,5), ylab = "y", xlab = "x",
     main = expression("varying"~beta[0]~"and"~beta[1]),
     xlim = c(-1,1.1),
     pch = 19, col = alpha(pal_dgp[1], 0.4), cex = 0.7)
abline(coef = c(-1,1), col = pal_dgp[1], lwd = 2)
abline(coef = c(1,-1), col = pal_dgp[2], lwd = 2)
points(x = d$x[d$z1>xi], y = d$y[d$z1>=xi], pch = 19, col = alpha(pal_dgp[2], 0.4), cex = 0.7)
legend("top", c(expression(z[1]<=xi), expression(z[1]>xi)), col = c(pal_dgp[1], pal_dgp[2]), pch = c(19), bty = "n", cex = 0.9)
mtext(expression(beta[0]==-delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 1.7)
mtext(expression(beta[1]==+delta), side = 4, col = pal_dgp[1], las = 1, line = 0.5, at = 0.7)
mtext(expression(beta[0]==+delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -0.7)
mtext(expression(beta[1]==-delta), side = 4, col = pal_dgp[2], las = 1, line = 0.5, at = -1.7)
@
\endminipage
\caption{\label{fig:dgp_stump}{Top left panel: True stump structure applied in the data generating process for the ``stump'' scenario with the true model parameters $\beta_0$
and $\beta_1$, either varying or being fixed at $0$ or $1$ respectively.
Top right and bottom panels: Bivariate plot of the response $Y$ on regressor $X$ 
with effect size $\delta = 1$ and for three variations: 
varying $\beta_0$ and fixed $\beta_1$ (top right),
fixed $\beta_0$ and varying $\beta_1$ (bottom left),
varying $\beta_0$ and varying $\beta_1$ (bottom right).}}
\end{figure}

<<true_tree_dgp, eval = TRUE, echo=FALSE, results=hide>>=
# d <- dgp_tree(nobs = 250)
set.seed(7)
d <- dgp_tree(nobs = 250, xi = 0, variation = "all", 
              changetype= "abrupt", sigma = 0.5)

@@

library("latticeExtra")
useOuterStrips(xyplot(prop_Tsplit ~ delta | variation + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = changetype == "abrupt", 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Selection probability of"~Z[1]),
                      par.settings = par.settings),

@@

<<3way_data, results=hide, echo=FALSE>>=
if(file.exists("../simulation/results/sim_3way.rda")){
  load("../simulation/results/sim_3way.rda")
} else {
  # for delta = 0.3 and xi = 0
  sim3way1 <- sim(scenario = "stump",
                  nobs = 250, nrep = 100, seed = 7,
                  delta = 0.3,
                  xi = 0, 
                  changetype = "abrupt",
                  variation = "all", 
                  return_matrices = TRUE,
                  test = c("guide_sum_1_cor",
                           "mfluc_resid_bin",
                           "ctree_resid_bin",
                           "mfluc_resid_cat",
                           "mfluc_resid",
                           "ctree_resid",
                           "guide_sum_12",
                           "mfluc_bin",
                           "ctree_bin",
                           "mfluc_cat",
                           "mfluc",
                           "ctree"))
  # prepare results for illustrations
  sim_3way1 <- prep_3way(sim_3way1)

  # for delta = 1 and xi = 0.8
  sim_3way2 <- sim(scenario = "stump",
                   nobs = 250, nrep = 100, seed = 7,
                   delta = 1,
                   xi = 0.8, 
                   changetype = "abrupt",
                   variation = "all",
                   return_matrices = TRUE,
                   test = c("guide_sum_1_cor",
                            "mfluc_resid_bin",
                            "ctree_resid_bin",
                            "mfluc_resid_cat",
                            "mfluc_resid",
                            "ctree_resid",
                            "guide_sum_12",
                            "mfluc_bin",
                            "ctree_bin",
                            "mfluc_cat",
                            "mfluc",
                            "ctree"))
  # prepare results for illustrations
  sim_3way2 <- prep_3way(sim_3way2)

  sim_3way1$xi <- 0
  sim_3way2$xi <- 0.8

  # combine results for illustration in two panels as presented in Section 5.2. (Figure 4)
  sim_3way <- rbind(sim_3way1, sim_3way2)
  sim_3way$xi <- factor(sim_3way$xi, levels = c(0,0.8), labels = c(0,0.8))

  save(sim_3way, file = "../simulation/results/sim_3way.rda")
}
@

\begin{figure}[t!]
\setkeys{Gin}{width=1\linewidth}
<<3way, fig=TRUE, echo=FALSE, height=4, width=8>>=
xyplot(pval_z1 ~ cat | xi, 
       groups = interaction(bin, res_scores), 
       data = sim_3way, 
       ylim = c(0,0.5),
       type = c("a"), 
       col = c(pal[1], pal[3], pal[1], pal[3]),
       lty = c(1,1,2,2),
       lwd = 2,
       scales = list(alternating=FALSE),
       xlab = "Categorization",
       ylab = expression("p-value of"~Z[1]),
       strip = strip.custom(factor.levels = c(expression(xi==0~"(50%)"~" and"~delta==0.3), 
                                              expression(xi==0.8~"(90%)"~" and"~delta==1)),
                            bg = "gray90"),
       key = list(text = list(c("Goodness of fit:", "  residuals", "  scores", "Dichotomization:", "  yes", "  no")),
                  columns = 2,
                  lines = list(lty = c(0,1,2,0,1,1), 
                               type = "l", lwd = 2,
                               col = c("black","black","black","black", pal[1], pal[3]))))

@
\caption{\label{fig:3way} Effects of the three building blocks \textit{residuals vs.\ scores},
\textit{dichotomization}, and \textit{categorization} in the ``stump'' scenario with the intercept
and the slope parameter both varying. All possible combinations have been 
evaluated in two different settings and their performances are compared based on the 
mean $p$-values corresponding to the true split variable $Z_1$. 
Left panel: true split point at the median $0$ and with
effect size $0.3$.
Right panel: true split point at the $90$\% quantile $0.8$ and with effect 
size $1$.}
\end{figure}

@@

<<tree_all_data, echo=FALSE, results=hide>>=
if(file.exists("../simulation/results/sim_tree.rda")){
  load("../simulation/results/sim_tree.rda")
} else {
  sim_tree <- simwrapper(scenario = "tree",
                         nobs = 250, nrep = 100, seed = 7, 
                         delta = seq(from = 0, to = 1, by = 0.1),
                         xi = c(0, 0.2, 0.5, 0.8), 
                         variation = "all", 
                         changetype = "abrupt",
                         compare_pruning = TRUE,
                         test = c("ctree", "mfluc",
                                  "guide_sum_12", 
                                  "guide_sum_1_cor"))

  # prepare data for illustration as presented in Section 5.3. (Figure 5)
  sim_tree <- sim_tree$res
  # extract data for xyplot comparing ARI and ARI_p
  sim_tree <- sim_tree[,c(1,2,5,6,7,10,13)]
  # extended data set with post-pruning results being represented as separated tests
  sim_tree_ex <- sim_tree[,c(1,2,3,4,6)]
  sim_tree_ex <- rbind(sim_tree_ex, sim_tree_ex)

  sim_tree_ex$test <- factor(sim_tree_ex$test,
                             levels = c("ctree","mfluc",
                                        "guide_sum_12", "guide_sum_1_cor"),
                             labels = c("CTree","MOB",
                                        "GUIDE+scores", "GUIDE"))

  sim_tree_ex$nrsubgr[(length(sim_tree$nrsubgr)+1) : (2*length(sim_tree$nrsubgr))] <- sim_tree$nrsubgr_p
  sim_tree_ex$ari[(length(sim_tree$ari)+1) : (2*length(sim_tree$ari))] <- sim_tree$ari_p
  sim_tree_ex$pruning <- c(rep("pre", NROW(sim_tree)), rep("post", NROW(sim_tree)))
  sim_tree <- sim_tree_ex
  rm(sim_tree_ex)
  save(sim_tree, file = "../simulation/results/sim_tree.rda")  
}

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<tree_all, fig=TRUE, echo=FALSE, height=5, width=8>>=
library("latticeExtra")
par.settings <- list(superpose.symbol = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], 
                                             fill = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]), 
                     superpose.line = list(col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")]))
useOuterStrips(xyplot(ari ~ delta | xi + pruning, groups = ~ test, 
                      data = sim_tree, 
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      key =  list(text = list(levels(sim_tree$test)[1:4]),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(4,2),
                      scales=list(x=list(at=seq(1,11,2))),
                      xlab = expression(delta),
                      ylab = "Adjusted Rand Index",
                      #index.cond = list(c(2,3,1),c(2,1)),
                      par.settings = par.settings),
               strip = strip.custom(factor.levels = c(expression(xi==0), 
                                                      expression(xi==0.2), 
                                                      expression(xi==0.5), 
                                                      expression(xi==0.8)),
                                    bg = "gray90"),
               strip.left=strip.custom(factor.levels = c("Post-pruning", "Pre-pruning"),
                                       bg = "gray90"))

@
\caption{\label{fig:tree_all}
Adjusted Rand index (ARI) for the testing strategies CTree, MOB, GUIDE, and GUIDE+scores in the
``tree'' scenario, once applying pre-pruning (top row) and once post-pruning (bottom row). 
The ARI is evaluated over
increasing effect size $\delta$ with 100 replications of 250 observations per step.
Each panel represents a different value of the true split point $\xi$ ($0$, $0.2$, $0.5$ and 
$0.8$ from left to right, corresponding to the $50$\%, $60$\%, $75$\%, and $90$\% quantile).
}
\end{figure}

@@

useOuterStrips(xyplot(prop_z1 ~ delta | variation + xi, groups = ~ test, 
                      data = subdataxi, 
                      subset = changetype== "abrupt",
                      type = "l", lty = c(1,1,1,1),
                      lwd = 2,
                      scales=list(x=list(at=seq(1,11,2))),
                      key =  list(text = list(levels(subdataxi$test)),
                                  lines = list(lty = c(1,1,1,1), type = "l", pch = 1, cex = 0.8,
                                               col = pal[c("CTree", "MOB", "GUIDE+scores", "GUIDE")], lwd = 2)),
                      layout= c(3,2),
                      index.cond = list(c(2,3,1),c(2,1)),
                      xlab = expression(delta),
                      ylab = expression("Prop. of"~Z[1]~"having lowest p-value"),
                      par.settings = par.settings),

@@

\begin{figure}[t!]
\setkeys{Gin}{width=\linewidth}
<<contbeta, fig=TRUE, echo=FALSE, height=4, width=8>>=
subdata <- subset(sim_stump, test %in% c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"))
subdata$test <- factor(subdata$test, 
                       levels = c("ctree", "ctree_max","mfluc", "guide_sum_1_cor","guide_sum_12"),
                       labels = c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores"))
par.settings <- list(superpose.symbol = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
                                             fill = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]), 
                     superpose.line = list(col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")]))

xyplot(prop_Tsplit ~ delta | variation, groups = ~ test, 
       data = subdata, 
       subset = changetype == "continuous", 
       scales=list(x=list(at=seq(1,11,2))),
       type = "l", 
       layout= c(3,1),
       lwd = 2,
       key =  list(text = list(levels(subdata$test)),
                   lines = list(lty = c(1,1,1,1,1), 
                                type = "l", 
                                col = pal[c("CTree","CTree+max","MOB", "GUIDE","GUIDE+scores")], 
                                lwd = 2)), 
       xlab = expression(delta),
       ylab = expression("Selection probability of"~Z[1]),
       index.cond = list(c(2,3,1)),
       strip = strip.custom(factor.levels = c(expression("varying"~beta[0]~" and "~beta[1]),
                                              expression("varying"~beta[0]), 
                                              expression("varying"~beta[1])),
                            bg = "gray90"),
       par.settings = par.settings)

@
\caption{\label{fig:stump_contbeta}
Selection probability of the true split variable $Z_1$ for testing strategies CTree, CTree+max,
MOB, GUIDE, and GUIDE+scores for continuous parameter functions $\beta_0(Z_1)$ and $\beta_1(Z_1)$.
Probabilities are estimated over increasing effect size $\delta$
with 100 replications of 250 observations per step. 
Each panel represents one variation for the coefficients: 
varying intercept $\beta_0$ (left), varying slope parameter $\beta_1$ (middle) and 
both coefficients varying (right).}
\end{figure}
